### 2025-03-13 paper/presenations/slides.qmd
some work on cutting content

---
title: Counterfactual Training
subtitle: Update Meeting Dec 2024
author: 
  - name: Patrick Altmeyer
    url: https://www.patalt.org/
  - name: Arie van Deursen
  - name: Cynthia C. S. Liem
institute: Delft University of Technology
date: today
format: 
  beamer:
    theme: Boadilla
  tudelft-revealjs:
    theme: custom.scss
    self-contained: true
    smaller: false
    scrollable: true
    preview-links: auto
    slide-number: true
    transition: slide
    background-transition: fade
    fig-align: center
    html-math-method: mathjax
    include-in-header:
      - text: |
          <script>
          MathJax = {
            options: {
              menuOptions: {
                settings: {
                  assistiveMml: false
                }
              }
            }
          };
          </script>
crossref: 
  prp-title: RQ
  prp-prefix: RQ
classoption: "notheorems"
---

@@

Counterfactual Training (CT) combines ideas from Energy-Based Models and Adversarial Training:

>>>>

---
title: counterfactual training
subtitle: Update Meeting Dec 2024
author: 
  - name: Patrick Altmeyer
    url: https://www.patalt.org/
  - name: Arie van Deursen
  - name: Cynthia C. S. Liem
institute: Delft University of Technology
date: today
format: 
  beamer:
    theme: Boadilla
  tudelft-revealjs:
    theme: custom.scss
    self-contained: true
    smaller: false
    scrollable: true
    preview-links: auto
    slide-number: true
    transition: slide
    background-transition: fade
    fig-align: center
    html-math-method: mathjax
    include-in-header:
      - text: |
          <script>
          MathJax = {
            options: {
              menuOptions: {
                settings: {
                  assistiveMml: false
                }
              }
            }
          };
          </script>
crossref: 
  prp-title: RQ
  prp-prefix: RQ
classoption: "notheorems"
---

@@

counterfactual training (CT) combines ideas from Energy-Based Models and Adversarial Training:

### 2025-03-13 paper/quarto_ecml/paper.tex
some work on cutting content

\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage[misc]{ifsym}
\newcommand{\corr}{(\Letter)}
\usepackage{placeins}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Fig.}
\else
  \newcommand\figurename{Fig.}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

@@

\begin{abstract}
We propose a novel training regime termed Counterfactual Training that
leverages counterfactual explanations to increase the explanatory
capacity of models. Counterfactual explanations have emerged as a
popular post-hoc explanation method for opaque machine learning models:
they inform how factual inputs would need to change in order for a model
to produce some desired output. To be useful in real-word
decision-making systems, counterfactuals should be plausible with
respect to the underlying data and actionable with respect to the
stakeholder requirements. Much existing research has therefore focused
on developing post-hoc methods to generate counterfactuals that meet
these desiderata. In this work, we instead hold models directly
accountable for the desired end goal: Counterfactual Training employs
counterfactuals ad-hoc during the training phase to minimize the
divergence between learned representations and plausible, actionable
explanations. We demonstrate empirically and theoretically that our
proposed method facilitates training models that deliver inherently
desirable explanations while maintaining high predictive performance.

@@

Part of the problem is that the high degrees of freedom provide room for
many solutions that are locally optimal with respect to narrow
objectives \cite{wilson2020case}.\footnote{We follow the standard ML
  convention, where ``degrees of freedom'' refer to the number of
  parameters estimated from data.} Indeed, recent work on the so called
``lottery ticket hypothesis'' suggests that modern neural networks can
be pruned by up to 90\% while preserving their predictive performance
\cite{frankle2018lottery} and generalizability \cite{morcos2019success}.
Similarly, \cite{zhang2021understanding} showed that state-of-the-art
neural networks are so expressive that they can fit randomly labeled
data. Thus, looking at the predictive performance alone, the solutions
may seem to provide compelling explanations for the data, when in fact
they are based on purely associative, semantically meaningless patterns.
This poses two related challenges. Firstly, there is no dependable way
to verify if such complex representations correspond to meaningful and
plausible explanations. Secondly, even if we could resolve the first
challenge, it remains undecided how to ensure that models can
\emph{only} learn valuable explanations.

@@

\textbf{Specifically, we introduce counterfactual training}: a novel
training regime explicitly meant to align learned representations with
plausible explanations that comply with user requirements. Our
contributions are as follows:

\begin{itemize}
\tightlist
\item
  We present a novel methodological framework that leverages adversarial
  examples and faithful counterfactual explanations during the training
  phase to improve the explanatory capacity and robustness of machine
  learning models (Section~\ref{sec-method}).\\
\item
  We propose a method to enforce global actionability constraints by
  preventing models from assigning importance to immutable features,
  i.e., ones over which decision subjects have no control
  (Section~\ref{sec-method}).\\
\item
  Through extensive experiments we demonstrate that counterfactual
  training promotes explainability while preserving high predictive
  performance. We run ablation studies and grid searches to understand
  how the underlying model components and hyperparameters affect
  outcomes. (Section~\ref{sec-experiments}).
\end{itemize}

Despite some limitations discussed in Section~\ref{sec-discussion}, we
conclude in Section~\ref{sec-conclusion} that counterfactual training
provides a useful framework for researchers and practitioners interested
in making opaque models more trustworthy. We also believe that this work
serves as an opportunity for XAI researchers to re-evaluate the trend of
improving XAI tools without improving the underlying models.

@@

Once again it helps to look at these findings through the lens of
representation learning with high degrees of freedom. Deep ensembles are
approximate Bayesian model averages, which are most called for when
models are underspecified by the available data \cite{wilson2020case}.
Averaging across solutions mitigates the aforementioned risk of relying
on a single locally optimal representations that corresponds to
semantically meaningless explanations for the data. Previous work by
\cite{schut2021generating} similarly found that generating plausible
(``interpretable'') counterfactual explanations is almost trivial for
deep ensembles that have also undergone adversarial training. The case
for JEMs is even clearer: they involve a hybrid objective that induces
both high predictive performance and generative capacity
\cite{grathwohl2020your}. This is closely related to the idea of
aligning models with plausible explanations and has inspired our
proposed counterfactual training objective, as we explain in
Section~\ref{sec-method}.

@@

Counterfactual training (CT) combines ideas from adversarial training,
energy-based modelling and counterfactuals explanations with the
explicit goal of aligning representations with plausible explanations
that comply with user requirements. In the context of CEs, plausibility
has broadly been defined as the degree to which counterfactuals comply
with the underlying data-generating process
\cite{poyiadzi2020face,guidotti2022counterfactual,altmeyer2024faithful}.
Plausibility is a necessary but insufficient condition for using CEs to
provide algorithmic recourse (AR) to individuals (negatively) affected
by opaque models. For AR recommendations to be actionable, they need to
not only result in plausible counterfactuals but also be attainable. A
plausible CE for a rejected 20-year-old loan applicant, for example,
might reveal that their application would have been accepted, if only
they were 20 years older. Ignoring all other features, this would comply
with the definition of plausibility if 40-year-old individuals were in
fact more credit-worthy on average than young adults. But of course this
CE does not qualify for providing actionable recourse to the applicant
since \emph{age} is not a (directly) mutable feature. CT aims to improve
model explainability by aligning models with counterfactuals that meet
both desiderata: plausibility and actionability. Formally, we define
explainability as follows:

@@

\begin{equation}\phantomsection\label{eq-obj}{
\begin{split}
\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x},\mathbf{x}_T^\prime,y;\theta) &+ \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y}) \\ &+ \lambda_{\text{reg}}\text{ridge}(\mathbf{x},\mathbf{x}_T^\prime,y;\theta)
\end{split}
}\end{equation}

where \(\text{yloss}(\cdot)\) is a classification loss that induces
discriminative performance (e.g., cross-entropy). The second and third
terms in Equation~\ref{eq-obj} are explained in detail below. For now,
they can be sufficiently described as inducing explainability directly
and indirectly by penalizing: (1) the contrastive divergence,
\(\text{div}(\cdot)\), between mature counterfactuals
\(\mathbf{x}_T^\prime\) and observed samples \(x\) and, (2) the
adversarial loss, \(\text{advloss}(.)\), with respect to nascent
counterfactuals \(\mathbf{x}_{t\leq T}^\prime\). Finally,
\(\text{ridge}(\cdot)\) denotes a Ridge penalty (\(\ell_2\)-norm) that
regularizes the magnitude of the energy terms involved in
\(\text{div}(\cdot)\) \cite{du2019implicit}. The trade-off between the
components can be governed by adjusting the strengths of the penalties
\(\lambda_{\text{div}}\), \(\lambda_{\text{adv}}\) and
\(\lambda_{\text{reg}}\).

@@

\begin{equation}\phantomsection\label{eq-div}{
\text{div}(\mathbf{x},\mathbf{x}_T^\prime,y;\theta) = \mathcal{E}_\theta(\mathbf{x},y) - \mathcal{E}_\theta(\mathbf{x}_T^\prime,y)
}\end{equation}

where \(\mathcal{E}_\theta(\cdot)\) denotes the energy function. We set
\(\mathcal{E}_\theta(\mathbf{x},\mathbf{y})=-\mathbf{M}_\theta(\mathbf{x^+})[y^+]\)
where \(y^+\) denotes the index of the randomly drawn target class,
\(y^+ \sim p(y)\), and \(\mathbf{x^+}\) denotes an observed sample from
target domain: \(\mathbf{X}^+=\{\mathbf{x}:y=y^+\}\). Conditional on the
target class \(y^+\), \(\mathbf{x}_T^\prime\) denotes a mature
counterfactual for a randomly sampled factual from a non-target class
generated with a gradient-based CE generator for up to \(T\) iterations.
Mature counterfactuals are ones that have either reached convergence
wrt. the decision threshold \(\tau\) or exhausted \(T\).

@@

Since such actionability constraints are binding at test time, we should
also impose them when generating \(\mathbf{x}^\prime\) during each
training iteration to inform model representations. Through their effect
on \(\mathbf{x}^\prime\), both types of constraints influence model
outcomes via Equation~\ref{eq-div}. Here it is crucial that we avoid
penalizing implausibility that arises due to mutability constraints. For
any mutability-constrained feature \(d\) this can be achieved by
enforcing \(\mathbf{x}[d] - \mathbf{x}^\prime[d]:=0\) whenever
perturbing \(\mathbf{x}^\prime[d]\) in the direction of
\(\mathbf{x}[d]\) would violate mutability constraints. Specifically, we
set \(\mathbf{x}[d] := \mathbf{x}^\prime[d]\) if:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Feature \(d\) is strictly immutable in practice.
\item
  We have \(\mathbf{x}[d]>\mathbf{x}^\prime[d]\), but feature \(d\) can
  only be decreased in practice.
\item
  We have \(\mathbf{x}[d]<\mathbf{x}^\prime[d]\), but feature \(d\) can
  only be increased in practice.
\end{enumerate}

From a Bayesian perspective, setting
\(\mathbf{x}[d] := \mathbf{x}^\prime[d]\) can be understood as assuming
a point mass prior for \(p(\mathbf{x})\) with respect to feature \(d\).
Intuitively, we think of this simply in terms ignoring implausibility
costs with respect to immutable features, which effectively forces the
model to instead seek plausibility with respect to the remaining
features. This in turn results in lower overall sensitivity to immutable
features, which we demonstrate empirically for different classifiers in
Section~\ref{sec-experiments}. Under certain conditions, this results
holds theoretically:\footnote{For the proof, see the supplementary
  appendix.}

@@

In panel (a), we have trained our model conventionally, and we do not
impose mutability constraints at test time. The generated
counterfactuals are all valid, but not plausible: they do not comply
with the distribution of the factual samples in the target class to the
point where they are clearly distinguishable from the ground-truth data.
In panel (b), we have trained our model with counterfactual training,
once again without any mutability constraints. We observe that the
counterfactuals are highly plausible, meeting the first objective of
Def. \ref{def-explainability}.

In panel (c), we have used conventional training again, this time
imposing the mutability constraint on \emph{age} at test time.
Counterfactuals are valid but involve some substantial reductions in
\emph{debt} for some individuals (very young applicants). By comparison,
counterfactual paths are shorter on average in panel (d), where we have
used counterfactual training and protected the immutable feature as
described in Section~\ref{sec-constraints}. We observe that due to the
classifier's lower sensitivity to \emph{age}, recourse recommendations
with respect to \emph{debt} are much more homogenous and do not
disproportionately punish younger individuals. The counterfactuals are
also plausible with respect to the mutable feature. Thus, we consider
the model in panel (d) as the most explainable according to Def.
\ref{def-explainability}.

@@

We first address the direct extensions of the counterfactual training
approach in Section~\ref{sec-approach}. Then, we look at its limitations
and challenges in Section~\ref{sec-limitations}.

\subsection{Future research}\label{sec-approach}

@@

\subsection{Limitations and challenges}\label{sec-limitations}

@@

\textbf{\emph{Interventions on features may impact fairness
downstream.}} Related to the point above, we provide a tool that allows
practitioners to modify the sensitivity of a model with respect to
certain features, which may have implication for the fair and equitable
treatment of individuals subject to automated decisions. As protecting a
set of features leads the model to assign higher relative importance to
unprotected features, model owners could misuse our solution by
enforcing explanations based on features that are more difficult to
modify by some (group of) individuals. For example, consider again the
Adult dataset where features such as workclass or education may be more
difficult to change for underpriviledged groups. When applied
irresponsibly, counterfactual training could result in an unfairly
assigned burden of recourse \cite{sharma2020certifai}, threatening the
equality of opportunity in the system \cite{bell2024fairness} and
potentially reinforcing social segregation \cite{gao2023impact}. Still,
as the referenced publications indicate, such phenomena are not specific
to CT; all types of ADM solutions without strong external protections
have been recognized to promote harmful power dynamics
\cite{maas2023machine}.

@@

State-of-the-art machine learning models are prone to learning complex
representations that cannot be interpreted by humans. Although post-hoc
explainability approaches have attracted major research interest, these
cannot guarantee that the explanations agree with the opaque model's
learned representation of data. As a step towards addressing this
challenge, we introduced counterfactual training, a novel training
regime that incentivizes highly-explainable models. Our approach leads
to explanations that are both plausible---compliant with the underlying
data-generating process---and actionable---compliant with user-specified
mutability constraints---and thus meaningful to their recipients.
Through extensive experiments we demonstrate that counterfactual
training satisfies its objectives while preserving the predictive
performance of the trained models. We also find that our approach can be
used to fine-tune conventionally-trained models and achieve similar
gains in explainability. Finally, this work showcases that it is
practical to improve models \emph{and} their explanations at the same
time.

>>>>

\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage[misc]{ifsym}
\newcommand{\corr}{(\Letter)}
\usepackage{placeins}
\makeatletter

\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Fig.}
\else
  \newcommand\figurename{Fig.}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter

\makeatother

@@

\begin{abstract}
We propose a novel training regime termed counterfactual training that
leverages counterfactual explanations to increase the explanatory
capacity of models. Counterfactual explanations have emerged as a
popular post-hoc explanation method for opaque machine learning models:
they inform how factual inputs would need to change in order for a model
to produce some desired output. To be useful in real-word
decision-making systems, counterfactuals should be plausible with
respect to the underlying data and actionable with respect to the
stakeholder requirements. Much existing research has therefore focused
on developing post-hoc methods to generate counterfactuals that meet
these desiderata. In this work, we instead hold models directly
accountable for the desired end goal: counterfactual training employs
counterfactuals ad-hoc during the training phase to minimize the
divergence between learned representations and plausible, actionable
explanations. We demonstrate empirically and theoretically that our
proposed method facilitates training models that deliver inherently
desirable explanations while maintaining high predictive performance.

@@

Part of the problem is that the high degrees of freedom provide room for
many solutions that are locally optimal with respect to narrow
objectives \cite{wilson2020case}.\footnote{We follow the standard ML
  convention, where ``degrees of freedom'' refer to the number of
  parameters estimated from data.} Indeed, recent work on the so-called
``lottery ticket hypothesis'' suggests that modern neural networks can
be pruned by up to 90\% while preserving their predictive performance
\cite{frankle2018lottery} and generalizability \cite{morcos2019success}.
Similarly, \cite{zhang2021understanding} showed that state-of-the-art
neural networks are so expressive that they can fit randomly labeled
data. Thus, looking at the predictive performance alone, the solutions
may seem to provide compelling explanations for the data, when in fact
they are based on purely associative, semantically meaningless patterns.
This poses two related challenges. Firstly, there is no dependable way
to verify if such complex representations correspond to meaningful and
plausible explanations. Secondly, even if we could resolve the first
challenge, it remains undecided how to ensure that models can
\emph{only} learn valuable explanations.

@@

\textbf{Specifically, we introduce counterfactual training (CT)}: a
novel training regime explicitly meant to align learned representations
with plausible explanations that comply with user requirements. The
remainder of this paper is structured as follows. Section~\ref{sec-lit}
presents related work, focusing in particular on the link between
adversarial examples and counterfactual explanations. Then follow our
main contributions:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  In Section~\ref{sec-method}, we introduce our methodological framework
  and show theoretically that it can be used to enforce global
  actionability constraints.\\
\item
  Through extensive experiments we demonstrate that CT substantially
  improves explainability without sacrificing predictive performance
  (Section~\ref{sec-experiments}).
\end{enumerate}

We discuss future research and challenges in
Section~\ref{sec-discussion} and conclude in
Section~\ref{sec-conclusion} that CT is a promising new approach towards
making opaque models more trustworthy.

@@

Once again it helps to look at these findings through the lens of
representation learning with high degrees of freedom. Deep ensembles are
approximate Bayesian model averages, which are most called for when
models are underspecified by the available data \cite{wilson2020case}.
Averaging across solutions mitigates the aforementioned risk of relying
on a single locally optimal representations that corresponds to
semantically meaningless explanations for the data. Previous work by
\cite{schut2021generating} similarly found that generating plausible
(``interpretable'') counterfactual explanations is almost trivial for
deep ensembles that have also undergone adversarial training. The case
for JEMs is even clearer: they involve a hybrid objective that induces
both high predictive performance and generative capacity
\cite{grathwohl2020your}. This is closely related to the idea of
aligning models with plausible explanations and has inspired our
proposed CT objective, as we explain in Section~\ref{sec-method}.

@@

Counterfactual training combines ideas from adversarial training,
energy-based modelling and counterfactuals explanations with the
explicit goal of aligning representations with plausible explanations
that comply with user requirements. In the context of CEs, plausibility
has broadly been defined as the degree to which counterfactuals comply
with the underlying data-generating process
\cite{poyiadzi2020face,guidotti2022counterfactual,altmeyer2024faithful}.
Plausibility is a necessary but insufficient condition for using CEs to
provide algorithmic recourse (AR) to individuals (negatively) affected
by opaque models. For AR recommendations to be actionable, they need to
not only result in plausible counterfactuals but also be attainable. A
plausible CE for a rejected 20-year-old loan applicant, for example,
might reveal that their application would have been accepted, if only
they were 20 years older. Ignoring all other features, this would comply
with the definition of plausibility if 40-year-old individuals were in
fact more credit-worthy on average than young adults. But of course this
CE does not qualify for providing actionable recourse to the applicant
since \emph{age} is not a (directly) mutable feature. CT aims to improve
model explainability by aligning models with counterfactuals that meet
both desiderata: plausibility and actionability. Formally, we define
explainability as follows:

@@

\begin{equation}\phantomsection\label{eq-obj}{
\begin{split}
\min_\theta \text{yloss}(\mathbf{M}_\theta(\mathbf{x}),\mathbf{y}) + \lambda_{\text{div}} \text{div}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta) &+ \lambda_{\text{adv}} \text{advloss}(\mathbf{M}_\theta(\mathbf{x}_{t\leq T}^\prime),\mathbf{y}) \\ &+ \lambda_{\text{reg}}\text{ridge}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta)
\end{split}
}\end{equation}

where \(\text{yloss}(\cdot)\) is a classification loss that induces
discriminative performance (e.g., cross-entropy). The second and third
terms in Equation~\ref{eq-obj} are explained in detail below. For now,
they can be sufficiently described as inducing explainability directly
and indirectly by penalizing: (1) the contrastive divergence,
\(\text{div}(\cdot)\), between mature counterfactuals
\(\mathbf{x}_T^\prime\) and observed samples
\(\mathbf{x}^+\in\mathcal{X}^+=\{\mathbf{x}:y=y^+\}\) in the target
class \(y^+\), and, (2) the adversarial loss, \(\text{advloss}(.)\),
with respect to nascent counterfactuals \(\mathbf{x}_{t\leq T}^\prime\).
Finally, \(\text{ridge}(\cdot)\) denotes a Ridge penalty
(\(\ell_2\)-norm) that regularizes the magnitude of the energy terms
involved in \(\text{div}(\cdot)\) \cite{du2019implicit}. The trade-off
between the components can be governed by adjusting the strengths of the
penalties \(\lambda_{\text{div}}\), \(\lambda_{\text{adv}}\) and
\(\lambda_{\text{reg}}\).

@@

\begin{equation}\phantomsection\label{eq-div}{
\text{div}(\mathbf{x}^+,\mathbf{x}_T^\prime,y;\theta) = \mathcal{E}_\theta(\mathbf{x}^+,y) - \mathcal{E}_\theta(\mathbf{x}_T^\prime,y)
}\end{equation}

where \(\mathcal{E}_\theta(\cdot)\) denotes the energy function. We set
\(\mathcal{E}_\theta(\mathbf{x},y)=-\mathbf{M}_\theta(\mathbf{x})[y^+]\)
where \(y^+\) denotes the index of the randomly drawn target class,
\(y^+ \sim p(y)\). Conditional on the target class \(y^+\),
\(\mathbf{x}_T^\prime\) denotes a mature counterfactual for a randomly
sampled factual from a non-target class generated with a gradient-based
CE generator for up to \(T\) iterations. Mature counterfactuals are ones
that have either reached convergence wrt. the decision threshold
\(\tau\) or exhausted \(T\).

@@

Since such actionability constraints are binding at test time, we should
also impose them when generating \(\mathbf{x}^\prime\) during each
training iteration to inform model representations. Through their effect
on \(\mathbf{x}^\prime\), both types of constraints influence model
outcomes via Equation~\ref{eq-div}. Here it is crucial that we avoid
penalizing implausibility that arises due to mutability constraints. For
any mutability-constrained feature \(d\) this can be achieved by
enforcing \(\mathbf{x}^+[d] - \mathbf{x}^\prime[d]:=0\) whenever
perturbing \(\mathbf{x}^\prime[d]\) in the direction of
\(\mathbf{x}^+[d]\) would violate mutability constraints. Specifically,
we set \(\mathbf{x}^+[d] := \mathbf{x}^\prime[d]\) if:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Feature \(d\) is strictly immutable in practice.
\item
  We have \(\mathbf{x}^+[d]>\mathbf{x}^\prime[d]\), but feature \(d\)
  can only be decreased in practice.
\item
  We have \(\mathbf{x}^+[d]<\mathbf{x}^\prime[d]\), but feature \(d\)
  can only be increased in practice.
\end{enumerate}

From a Bayesian perspective, setting
\(\mathbf{x}^+[d] := \mathbf{x}^\prime[d]\) can be understood as
assuming a point mass prior for \(p(\mathbf{x}^+)\) with respect to
feature \(d\). Intuitively, we think of this simply in terms ignoring
implausibility costs with respect to immutable features, which
effectively forces the model to instead seek plausibility with respect
to the remaining features. This in turn results in lower overall
sensitivity to immutable features, which we demonstrate empirically for
different classifiers in Section~\ref{sec-experiments}. Under certain
conditions, this results holds theoretically:\footnote{For the proof,
  see the supplementary appendix.}

@@

In panel (a), we have trained our model conventionally, and we do not
impose mutability constraints at test time. The generated
counterfactuals are all valid, but not plausible: they do not comply
with the distribution of the factual samples in the target class to the
point where they are clearly distinguishable from the ground-truth data.
In panel (b), we have trained our model with CT, once again without any
mutability constraints. We observe that the counterfactuals are highly
plausible, meeting the first objective of Def. \ref{def-explainability}.

In panel (c), we have used conventional training again, this time
imposing the mutability constraint on \emph{age} at test time.
Counterfactuals are valid but involve some substantial reductions in
\emph{debt} for some individuals (very young applicants). By comparison,
counterfactual paths are shorter on average in panel (d), where we have
used CT and protected the immutable feature as described in
Section~\ref{sec-constraints}. We observe that due to the classifier's
lower sensitivity to \emph{age}, recourse recommendations with respect
to \emph{debt} are much more homogenous and do not disproportionately
punish younger individuals. The counterfactuals are also plausible with
respect to the mutable feature. Thus, we consider the model in panel (d)
as the most explainable according to Def. \ref{def-explainability}.

@@

We first address the direct extensions of CT in
Section~\ref{sec-approach}. Then, we look at its limitations and
challenges in Section~\ref{sec-limitations}.

\subsection{Future Research}\label{sec-approach}

@@

\subsection{Limitations and Challenges}\label{sec-limitations}

@@

\textbf{\emph{Interventions on features may impact fairness
downstream.}} Related to the point above, we provide a tool that allows
practitioners to modify the sensitivity of a model with respect to
certain features, which may have implication for the fair and equitable
treatment of individuals subject to automated decisions. As protecting a
set of features leads the model to assign higher relative importance to
unprotected features, model owners could misuse our solution by
enforcing explanations based on features that are more difficult to
modify by some (group of) individuals. For example, consider again the
Adult dataset where features such as workclass or education may be more
difficult to change for underpriviledged groups. When applied
irresponsibly, CT could result in an unfairly assigned burden of
recourse \cite{sharma2020certifai}, threatening the equality of
opportunity in the system \cite{bell2024fairness} and potentially
reinforcing social segregation \cite{gao2023impact}. Still, as the
referenced publications indicate, such phenomena are not specific to CT;
all types of ADM solutions without strong external protections have been
recognized to promote harmful power dynamics \cite{maas2023machine}.

@@

State-of-the-art machine learning models are prone to learning complex
representations that cannot be interpreted by humans. Although post-hoc
explainability approaches have attracted major research interest, these
cannot guarantee that the explanations agree with the opaque model's
learned representation of data. As a step towards addressing this
challenge, we introduced counterfactual training, a novel training
regime that incentivizes highly-explainable models. Our approach leads
to explanations that are both plausible---compliant with the underlying
data-generating process---and actionable---compliant with user-specified
mutability constraints---and thus meaningful to their recipients.
Through extensive experiments we demonstrate that CT satisfies its
objectives while preserving the predictive performance of the trained
models. We also find that our approach can be used to fine-tune
conventionally-trained models and achieve similar gains in
explainability. Finally, this work showcases that it is practical to
improve models \emph{and} their explanations at the same time.

### 2025-03-13 paper/sections/abstract.qmd
some work on cutting content

We propose a novel training regime termed Counterfactual Training that leverages counterfactual explanations to increase the explanatory capacity of models. Counterfactual explanations have emerged as a popular post-hoc explanation method for opaque machine learning models: they inform how factual inputs would need to change in order for a model to produce some desired output. To be useful in real-word decision-making systems, counterfactuals should be plausible with respect to the underlying data and actionable with respect to the stakeholder requirements. Much existing research has therefore focused on developing post-hoc methods to generate counterfactuals that meet these desiderata. In this work, we instead hold models directly accountable for the desired end goal: Counterfactual Training employs counterfactuals ad-hoc during the training phase to minimize the divergence between learned representations and plausible, actionable explanations. We demonstrate empirically and theoretically that our proposed method facilitates training models that deliver inherently desirable explanations while maintaining high predictive performance.

>>>>

We propose a novel training regime termed counterfactual training that leverages counterfactual explanations to increase the explanatory capacity of models. Counterfactual explanations have emerged as a popular post-hoc explanation method for opaque machine learning models: they inform how factual inputs would need to change in order for a model to produce some desired output. To be useful in real-word decision-making systems, counterfactuals should be plausible with respect to the underlying data and actionable with respect to the stakeholder requirements. Much existing research has therefore focused on developing post-hoc methods to generate counterfactuals that meet these desiderata. In this work, we instead hold models directly accountable for the desired end goal: counterfactual training employs counterfactuals ad-hoc during the training phase to minimize the divergence between learned representations and plausible, actionable explanations. We demonstrate empirically and theoretically that our proposed method facilitates training models that deliver inherently desirable explanations while maintaining high predictive performance.

### 2025-03-13 paper/sections/conclusion.qmd
some work on cutting content

# Conclusion {#sec-conclusion}
State-of-the-art machine learning models are prone to learning complex representations that cannot be interpreted by humans. Although post-hoc explainability approaches have attracted major research interest, these cannot guarantee that the explanations agree with the opaque model's learned representation of data. As a step towards addressing this challenge, we introduced counterfactual training, a novel training regime that incentivizes highly-explainable models. Our approach leads to explanations that are both plausible---compliant with the underlying data-generating process---and actionable---compliant with user-specified mutability constraints---and thus meaningful to their recipients. Through extensive experiments we demonstrate that counterfactual training satisfies its objectives while preserving the predictive performance of the trained models. We also find that our approach can be used to fine-tune conventionally-trained models and achieve similar gains in explainability. Finally, this work showcases that it is practical to improve models *and* their explanations at the same time.

>>>>

# Conclusion {#sec-conclusion}
State-of-the-art machine learning models are prone to learning complex representations that cannot be interpreted by humans. Although post-hoc explainability approaches have attracted major research interest, these cannot guarantee that the explanations agree with the opaque model's learned representation of data. As a step towards addressing this challenge, we introduced counterfactual training, a novel training regime that incentivizes highly-explainable models. Our approach leads to explanations that are both plausible---compliant with the underlying data-generating process---and actionable---compliant with user-specified mutability constraints---and thus meaningful to their recipients. Through extensive experiments we demonstrate that CT satisfies its objectives while preserving the predictive performance of the trained models. We also find that our approach can be used to fine-tune conventionally-trained models and achieve similar gains in explainability. Finally, this work showcases that it is practical to improve models *and* their explanations at the same time.

### 2025-03-13 paper/sections/discussion.qmd
some work on cutting content

We first address the direct extensions of the counterfactual training approach in @sec-approach. Then, we look at its limitations and challenges in @sec-limitations.

## Future research {#sec-approach}

@@

## Limitations and challenges {#sec-limitations}

@@

***Interventions on features may impact fairness downstream.***  Related to the point above, we provide a tool that allows practitioners to modify the sensitivity of a model with respect to certain features, which may have implication for the fair and equitable treatment of individuals subject to automated decisions. As protecting a set of features leads the model to assign higher relative importance to unprotected features, model owners could misuse our solution by enforcing explanations based on features that are more difficult to modify by some (group of) individuals. For example, consider again the Adult dataset where features such as workclass or education may be more difficult to change for underpriviledged groups. When applied irresponsibly, counterfactual training could result in an unfairly assigned burden of recourse [e.g., @sharma2020certifai], threatening the equality of opportunity in the system [@bell2024fairness] and potentially reinforcing social segregation [@gao2023impact]. Still, as the referenced publications indicate, such phenomena are not specific to CT; all types of ADM solutions without strong external protections have been recognized to promote harmful power dynamics [@maas2023machine].

>>>>

We first address the direct extensions of CT in @sec-approach. Then, we look at its limitations and challenges in @sec-limitations.

## Future Research {#sec-approach}

@@

## Limitations and Challenges {#sec-limitations}

@@

***Interventions on features may impact fairness downstream.***  Related to the point above, we provide a tool that allows practitioners to modify the sensitivity of a model with respect to certain features, which may have implication for the fair and equitable treatment of individuals subject to automated decisions. As protecting a set of features leads the model to assign higher relative importance to unprotected features, model owners could misuse our solution by enforcing explanations based on features that are more difficult to modify by some (group of) individuals. For example, consider again the Adult dataset where features such as workclass or education may be more difficult to change for underpriviledged groups. When applied irresponsibly, CT could result in an unfairly assigned burden of recourse [e.g., @sharma2020certifai], threatening the equality of opportunity in the system [@bell2024fairness] and potentially reinforcing social segregation [@gao2023impact]. Still, as the referenced publications indicate, such phenomena are not specific to CT; all types of ADM solutions without strong external protections have been recognized to promote harmful power dynamics [@maas2023machine].
